The \dealii{} Library, Version 9.4

### 2022-06-01 9.4/paper.tex
Update my affiliations + some text

 \author[6,7]{Martin~Kronbichler}
 \affil[6]{Institute for Computational Mechanics,
   Technical University of Munich,
   Boltzmannstr.~15, 85748 Garching, Germany.
   {\texttt{kronbichler/munch@lnm.mw.tum.de}}}
 \affil[7]{Department of Information Technology,
   Uppsala University,
   Box 337, 751\,05 Uppsala, Sweden.
   {\texttt{martin.kronbichler/simon.sticko@it.uu.se}}}

@@

\author[6,9]{Peter Munch}
 \affil[9]{Institute of Material Systems Modeling,
 Helmholtz-Zentrum Hereon,
 Max-Planck-Str. 1, 21502 Geesthacht, Germany.
   {\texttt{peter.muench@hereon.de}}}

@@

\author[7]{Simon~Sticko}

@@

Until now, distributed meshes were either partitioned statically (\texttt{parallel::\allowbreak fullydistributed::\allowbreak Triangulation}; short \texttt{p::f::T}) or they used a
fixed policy to partition the cells among processes
(\texttt{parallel::\allowbreak distributed::\allowbreak Triangulation}; short \texttt{p::d::T}). For the latter, we
use the Morton-order as the backend of \texttt{p::d::T}---\texttt{p4est}---does. A fixed Morton-order
might provide very good performance in terms in communication and
setup costs, but might be non-optimal when interfacing with other
libraries that have fixed partitioning, e.g., a Cartesian
partitioning, themselves.

@@

The setup process is visualized in Figure~\ref{fig:repartitioning}. At first, cells, incl. their
surrounding (ghost) cells and parent cells, are collected and 
sent to the new owner. On the
receiving side, the sets of all cells are combined and possible duplicates
are removed. This information is enough to construct a new triangulation.
For sending/receiving, we apply consensus-based algorithms~\cite{hoefler2010scalable}, which
we introduced into the library in release 9.2~\cite{dealII92} -- see
also Section~\ref{sec:CA}. Consensus-based
algorithms are used also in \cite{ibanez2016pumi} for repartitioning.

@@

In the matrix-free infrastructure, we have added a number of new features. The most
notable ones are:
\begin{itemize}
\item Improved support of Hessians: just like values and gradients, Hessians can be
now evaluated and integrated during matrix-free loops
both for cells and faces. This enables, e.g., to write
a matrix-free version of step-47, which solves the biharmonic equation with DG.
\item Cell-centric loops now also allow to access gradients and Hessians
of neighboring cells on faces. The major difficulty here was their reorientation
in the case of arbitrary orientated cells on unstructured meshes.
\item Users can now create their own cell batches, by providing \texttt{FEEvaluation::reinit()} a list of cell IDs. \texttt{FEEvaluation}
accesses the appropriate data and reshuffles mapping data accordingly on
the fly in order to enable vectorization over cells. The new feature is useful in many
contexts. Examples are application with sharp interfaces (e.g., two-phase flow
or shock capturing), where one needs to treat cells that are ``cut'' by
the interface in a special way. A challenge is that cell batches
might contain cut or not-cut cells, making vectorization potentially more complicated. One way to deal with such cell batches is to apply masks if
the code paths do not diverge too much
(a lot of functions of \texttt{FEEvaluation} allows this). Another way is to
categorize cells during \texttt{MatrixFree::reinit()} in such a way that mixed cell batches do not occur. However, calling  \texttt{MatrixFree::reinit()}
might be too expensive if a very dynamic system is given, which requires
recategorization in each time step. Doing this on the fly might be a cheep alternative, even if not computationally optimal during matrix-free loops.
\end{itemize}
Besides these new features, we improved the performance of 
hanging-node-constraint evaluation on the CPU. Instead of performing
quasi-dense matrix-vector multiplications~\cite{KronbichlerKormann2012}, we now use an
approach based on inplace interpolation and sumfactorization, similarly
as we already did in the GPU code~\cite{ljungkvist2017matrix}. In \cite{munch2022hn}, the algorithm is described 
and performance numbers are shown, indicating a reduction
of overhead of cells with hanging nodes by a factor of 10$\times$.

@@

In \cite{munch2022gc}, the performance of the local-smoothing and global-coarsening
infrastructure of \texttt{deal.II} was compared for locally refined meshes. The results indicate that
the local definition of multigrid levels might introduce load imbalances
in the case of local smoothing so that global coarsening is favorable despite
potentially more expensive intergrid transfers. In order, to judge the benefits
of one approach against the other, \texttt{deal.II} provides new functions
\texttt{workload\_imbalance()} and \texttt{vertical\_communication\_efficiency()} 
in the \texttt{MGTools} namespace for the  estimation of the workload imbalance and the
vertical communication efficiency purely based on the given mesh. The publication \cite{munch2022gc} also points out that not
all types of smoothers are applicable for global coarsening due to the
presence of hanging nodes, which is a motivation to add new smoother types
to \dealii in the future.

@@

Several classes have been added to the \texttt{NonMatching} namespace to enable the use of cut finite element methods\cite{burman_cutfem_2015}.
In the literature, these types of methods are also referred to as immersed, extended, or fictitious finite element methods.
Here, the domain, $\Omega$, is immersed in the background mesh, as illustrated in Figure~\ref{fig:immersed-domain}.
Often, one solves for the degrees of freedom of the smallest submesh which completely covers the domain.
The bilinear form in the weak form can, for example, look like
\begin{equation}
  a(u,v) = (\nabla u, \nabla v)_\Omega - (\partial_n u, v)_\Gamma + \ldots
\end{equation}
Thus, when assembling, we are on each cell
required to integrate over the part of the domain and the part of the boundary, $\Gamma = \partial \Omega$, that falls inside the cell:
$K\cap \Omega$ and $K \cap \Gamma$.
Many of the new classes assume that the domain is described by a level set function,
$\psi : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}
  \Omega = \{x \in \mathbb{R}^d : \psi(x)<0\},
  \qquad
  \Gamma = \{x \in \mathbb{R}^d : \psi(x) = 0\}.
\end{align}
\begin{itemize}
  \item The \texttt{MeshClassifier} class identifies how the active cells and faces are located relative to the zero contour of the level set function, as illustrated in Figure~\ref{fig:location-to-level-set}. Its \texttt{location\_to\_level\_set} function takes a cell/face and
        returns an enum, \texttt{ LocationToLevelSet}, which takes values \{\texttt{inside}, \texttt{outside}, \texttt{intersected}\}.
        These represent that $\psi$ is negative, positive, or varies in sign over the cell/face.
        That is, the cell/face is located inside/outside $\Omega$, or is intersected by $\Gamma$.
        This information is typically needed when choosing what element a cell of the \texttt{DoFHandler} should use.

@@

The 9.4 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_3_0_and_9_4_0.html}
{around 40 incompatible changes}; see \cite{changes94}. The majority of these changes
should not be visible to typical user codes; some remove previously
deprecated classes and functions; and the majority changes internal
interfaces that are not usually used in external
applications. That said, the following are worth mentioning since they
may have been more widely used:
\begin{itemize}
  \item In continuation of our attempt to merge the classes \texttt{DoFHandler} and \texttt{hp::DoFHandler}, we have removed the
  template parameter \texttt{DoFHandlerType} from number of classes and
  functions and replaced it by \texttt{dim}/\texttt{spacedim}. Affected
  classes are, e.g., \texttt{SolutionTransfer} and \texttt{DataOut}. 
  \item \texttt{FE\_RaviartThomasNodal} now uses a different polynomial space to allow
  for a simpler use for faces in non-standard orientation. The new polynomials
  are anisotropic tensor products of Lagrange polynomials on the points of the
  Gauss--Lobatto quadrature formula. This change leads to different entries, for example, in
  the matrices and constraints, but no change in accuracy should be expected as the resulting polynomial
  space spans the same polynomials.
  \item We made different changes in the repartitioning of \texttt{parallel::distributed::Triangulation}.
  Most notably, we have removed the default weight in order to ensure consistency with the rest of the
  library.
\end{itemize}

@@

M.~Kronbichler and P.~Munch were partially supported by the
Bayerisches Kompetenznetzwerk
f\"ur Technisch-Wissen\-schaft\-li\-ches Hoch- und H\"ochstleistungsrechnen
(KONWIHR) in the context of the projects
``Performance tuning of high-order discontinuous Galerkin solvers for
SuperMUC-NG'' and ``High-order matrix-free finite element implementations with
hybrid parallelization and improved data locality''.

>>>>

 \author[6,7]{Martin~Kronbichler}
 \affil[6]{Department of Information Technology,
   Uppsala University,
   Box 337, 751\,05 Uppsala, Sweden.
   {\texttt{martin.kronbichler/simon.sticko@it.uu.se}}}
 \affil[7]{Institute of Mathematics,
   University of Augsburg,
   Universit\"atsstr.~12a, 86159 Augsburg, Germany.
   {\texttt{martin.kronbichler@uni-a.de}}}

@@

\author[7,9]{Peter Munch}
 \affil[9]{Institute of Material Systems Modeling,
 Helmholtz-Zentrum Hereon,
 Max-Planck-Str. 1, 21502 Geesthacht, Germany.
   {\texttt{peter.muench@hereon.de}}}

@@

\author[6]{Simon~Sticko}

@@

Until now, distributed meshes were either partitioned statically (\texttt{parallel::\allowbreak fullydistributed::\allowbreak Triangulation}; short \texttt{p::f::T}) or they used a
fixed policy to partition the cells among processes
(\texttt{parallel::\allowbreak distributed::\allowbreak Triangulation}; short \texttt{p::d::T}). For the latter, we
use the Morton-order as the backend \texttt{p4est} does. A fixed Morton-order
might provide very good performance in terms in communication and
setup costs, but might be non-optimal when interfacing with other
libraries that have fixed partitioning, e.g., a Cartesian
partitioning, themselves.

@@

The setup process is visualized in Figure~\ref{fig:repartitioning}. At first, cells, including their
surrounding (ghost) cells and parent cells, are collected and
sent to the new owner. On the
receiving site, the sets of all cells are combined and possible duplicates
are removed. This information is enough to construct a new triangulation.
For sending/receiving, we apply consensus-based algorithms~\cite{hoefler2010scalable}, which
we introduced into the library in release 9.2~\cite{dealII92} -- see
also Section~\ref{sec:CA}. Consensus-based
algorithms are used also in \cite{ibanez2016pumi} for repartitioning.

@@

In the matrix-free infrastructure, we have added a number of new features. The most
notable ones are:
\begin{itemize}
\item Improved support of computations with Hessians of shape functions: just like values and gradients, Hessians can be
now evaluated and integrated during matrix-free loops
both for cells and faces. This enables, e.g., to write
a matrix-free version of step-47, which solves the biharmonic equation with DG.
\item Cell-centric loops now also allow to access gradients and Hessians
of neighboring cells on faces. The major difficulty here was their reorientation
in the case of arbitrary orientated cells on unstructured meshes.
\item Users can now create their own cell batches, by providing \texttt{FEEvaluation::reinit()} a list of cell IDs. \texttt{FEEvaluation}
accesses the appropriate data and reshuffles mapping data accordingly on
the fly in order to enable vectorization over cells. The new feature is useful in many
contexts. Examples are application with sharp interfaces (e.g., two-phase flow
or shock capturing), where one needs to treat cells that are ``cut'' by
the interface in a special way. A challenge is that cell batches
might contain cut or not-cut cells, making vectorization potentially more complicated. One way to deal with such cell batches is to apply masks if
the code paths do not diverge too much
(a lot of functions of \texttt{FEEvaluation} allows this). Another way is to
categorize cells during \texttt{MatrixFree::reinit()} in such a way that mixed cell batches do not occur. However, calling  \texttt{MatrixFree::reinit()}
might be too expensive if a very dynamic system is given, which requires
recategorization in each time step. Doing this on the fly might be a cheep alternative, even if not computationally optimal during matrix-free loops.
\item Initial support for H(div)-conforming elements with Piola transform, based on Raviart--Thomas finite element spaces with the updated class \texttt{FE\_RaviartThomasNodal}, has been added. The features are currently limited to meshes in standard orientations and affine geometries. Full support will be provided in the next release.
\item Selected matrix-free algorithms can now exploit additional data locality
  between the matrix-vector product and vector operations happening nearby in
  an algorithms. These have been added to both the
  \texttt{PreconditionChebyshev} class often used as a smoother in multigrid
  methods and the conjugate gradient implementation in \texttt{SolverCG}. To
  use these features, an operator needs to define a \texttt{vmult} operation,
  taking two additional \texttt{std::function} arguments. The first function
  defines what should be scheduled on the vector entries before the
  matrix-vector product touches them, and the second what happens
  afterwards. The new features also include a renumbering to maximize data
  locality. The theory is described in the
  contribution~\cite{kronbichler2022cg}.
\todo[inline]{We could also make this last a separate section or in the introduction, but it feels it fits well here.}
\end{itemize}
Besides these new features, we improved the performance of
hanging-node-constraint evaluation on the CPU. Instead of performing
quasi-dense matrix-vector multiplications~\cite{KronbichlerKormann2012}, we now use an
approach based on inplace interpolation and sumfactorization, similarly
as we already did in the GPU code~\cite{ljungkvist2017matrix}. In \cite{munch2022hn}, the algorithm is described
and performance numbers are shown, indicating a reduction
of overhead of cells with hanging nodes by a factor of 10$\times$.

@@

In \cite{munch2022gc}, the performance of the local-smoothing and global-coarsening
infrastructure of \texttt{deal.II} was compared for locally refined meshes. The results indicate that
the local definition of multigrid levels might introduce load imbalances
in the case of local smoothing so that global coarsening is favorable despite
potentially more expensive intergrid transfers. In order to judge the benefits
of one approach against the other, \texttt{deal.II} provides new functions
\texttt{workload\_imbalance()} and \texttt{vertical\_communication\_efficiency()}
in the \texttt{MGTools} namespace for the  estimation of the workload imbalance and the
vertical communication efficiency purely based on the given mesh. The publication \cite{munch2022gc} also points out that not
all types of smoothers are applicable for global coarsening due to the
presence of hanging nodes, which is a motivation to add new smoother types
to \dealii in the future.

@@

Several classes have been added to the \texttt{NonMatching} namespace to enable the use of cut finite element methods~\cite{burman_cutfem_2015}.
In the literature, these types of methods are also referred to as immersed, extended, or fictitious finite element methods.
Here, the domain, $\Omega$, is immersed in the background mesh, as illustrated in Figure~\ref{fig:immersed-domain}.
Often, one solves for the degrees of freedom of the smallest submesh which completely covers the domain.
The bilinear form in the weak form can, for example, look like
\begin{equation}
  a(u,v) = (\nabla u, \nabla v)_\Omega - (\partial_n u, v)_\Gamma + \ldots
\end{equation}
Thus, when assembling, we are on each cell
required to integrate over the part of the domain and the part of the boundary, $\Gamma = \partial \Omega$, that falls inside the cell:
$K\cap \Omega$ and $K \cap \Gamma$.
Many of the new classes assume that the domain is described by a level set function,
$\psi : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}
  \Omega = \{x \in \mathbb{R}^d : \psi(x)<0\},
  \qquad
  \Gamma = \{x \in \mathbb{R}^d : \psi(x) = 0\}.
\end{align}
\begin{itemize}
  \item The \texttt{MeshClassifier} class identifies how the active cells and faces are located relative to the zero contour of the level set function, as illustrated in Figure~\ref{fig:location-to-level-set}. Its \texttt{location\_to\_level\_set} function takes a cell/face and
        returns an enum, \texttt{LocationToLevelSet}, which takes values \{\texttt{inside}, \texttt{outside}, \texttt{intersected}\}.
        These indicate whether $\psi$ is negative, positive, or varies in sign over the cell/face.
        That is, the cell/face is located inside/outside $\Omega$, or is intersected by $\Gamma$.
        This information is typically needed when choosing what element a cell of the \texttt{DoFHandler} should use.

@@

The 9.4 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_3_0_and_9_4_0.html}
{around 40 incompatible changes}; see \cite{changes94}. The majority of these changes
should not be visible to typical user codes; some remove previously
deprecated classes and functions; and the majority changes internal
interfaces that are not usually used in external
applications. That said, the following are worth mentioning since they
may have been more widely used:
\begin{itemize}
  \item In continuation of our attempt to merge the classes \texttt{DoFHandler} and \texttt{hp::DoFHandler}, we have removed the
  template parameter \texttt{DoFHandlerType} from number of classes and
  functions and replaced it by \texttt{dim}/\texttt{spacedim}. Affected
  classes are, e.g., \texttt{SolutionTransfer} and \texttt{DataOut}.
  \item \texttt{FE\_RaviartThomasNodal} now uses a different polynomial space to allow
  for a simpler use for faces in non-standard orientation. The new polynomials
  are anisotropic tensor products of Lagrange polynomials on the points of the
  Gauss--Lobatto quadrature formula. This change leads to different entries, for example, in
  the matrices and constraints, but no change in accuracy should be expected as the resulting polynomial
  space spans the same polynomials.
  \item The class \texttt{MappingQ} now applies a high-order mapping to all cells, not just the cells near the boundary, functionality that was previously provided by the \texttt{MappingQGeneric} class. The latter has been marked as deprecated. In order to use different degrees for the mapping, users can use \texttt{hp::MappingCollection} or \texttt{MappingFEField} with appropriate definitions of function spaces.
  \item We made different changes in the repartitioning of \texttt{parallel::distributed::Triangulation}.
  Most notably, we have removed the default weight in order to ensure consistency with the rest of the
  library.
\end{itemize}

@@

M.~Kronbichler and P.~Munch were partially supported by the
Bayerisches Kompetenznetzwerk
f\"ur Technisch-Wissen\-schaft\-li\-ches Hoch- und H\"ochstleistungsrechnen
(KONWIHR) in the context of the projects
``High-order matrix-free finite element implementations with
hybrid parallelization and improved data locality'' and ``Fast and scalable finite element algorithms for coupled multiphysics problems and non-matching grids''.
